---
title: "Air Quality Data - Shawn, Jonas, and Walter"
# yaml-author
date: 2020-06-05
output: github_document
---
  
*Purpose*: Is there a correlation between covid-19 outcomes and air quality?

*Background*:

Air pollution is a mixture of solid and gas particles. Because air pollution may originate from different sources depending on location,  

Breathing in pollutants from the air is a large risk factor for lung related death. Studies have shown that long term exposure to polluted air can lead to irritation and damage throughout your respiratory pathway. Additionally, this damage is exacerbated in people with existing respiratory illnesses. Covid-19 is a novel coronavirus that emerged in late 2019 and has spread throughout the US and the world affecting millions. The virus and its effects are still being studied, but it has been shown to attack the lungs and cause problems with breathing. Due to the virus' impact on the respiratory system,  a study was conducted by Harvard researchers to analyze the connection between air pollution and Covid-19 mortality in the US. 

https://www.hsph.harvard.edu/news/hsph-in-the-news/air-pollution-linked-with-higher-covid-19-death-rates/
https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4776742/

*Notes*:
- I had trouble running the Case study in my browser. Note that the `who` dataset is loaded by the `tidyverse`. You can run the Case study locally if you need to!
- The case study uses `gather` instead of `pivot_longer`; feel free to use `pivot_longer` in place.

```{r setup, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

We load the libraries here: 

```{r library}
library(tidyverse)
library(tidyr)
```
We load the our datasets here: 

```{r load data}

#Loading EPA Air Quality Data
df_airquality2020<- read.csv("./data/annual_aqi_by_county_2020.csv")
df_airquality2020

#Load NYTimes Data - using code from c06
url_counties <- "https://raw.githubusercontent.com/nytimes/covid-19-data/master/live/us-counties.csv"
filename_nyt <- "./data/nyt_counties.csv"

## Download the data locally
curl::curl_download(
        url_counties,
        destfile = filename_nyt
      )

## Loads the downloaded csv
df_covid <- read_csv(filename_nyt)
df_covid
```

```{r wrangling data}
names(df_covid)[2] <- "County"
names(df_covid)[3] <- "State"
df_covid_air <- inner_join(df_covid, df_airquality2020, by = c("County", "State"))
df_covid_air #This is the joined version of the COVID-19 and EPA data set

```









