---
title: "personal_income"
author: "Jonas"
date: "10/20/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r setup}
library(tidyverse)

```


## R Markdown



```{r load data}
url_counties <- "https://raw.githubusercontent.com/nytimes/covid-19-data/master/live/us-counties.csv"
df_inc_dirty <- read.csv("data/Personal_Income.csv")
filename_nyt <- "./data/nyt_counties.csv"

## Download the data locally
curl::curl_download(
        url_counties,
        destfile = filename_nyt
      )

## Loads the downloaded csv
df_covid <- read_csv(filename_nyt)
```

## Including Plots

You can also embed plots, for example:

```{r glimpse}
df_covid %>% glimpse()
df_inc_dirty %>% glimpse()

```

```{r}
df_inc <- df_inc_dirty %>% 
  rename("Type" = X) %>%
  pivot_longer(-"Type", names_to = "Month")


df_inc
```


```{r}
df_inc %>%
  filter(Type == "Personal income") %>%
  ggplot() +
  geom_point(aes(Month, value))
```
a. What’s your question?

How does covid-19 cases by state relate to personal income?
or
How does a state's response to covid-19 affect personal income?

b. What data do you plan to use to answer this question?

https://www.bea.gov/data/income-saving/personal-income
https://www.bea.gov/data/income-saving/personal-income-by-state

c. What challenges do you anticipate encountering?

We only have data on personal income by state for q1 and q2, so we may run out of things to relate 

d. What level of complexity of final product are you aiming for?

1 being MVP, 5 being above and beyond 
2.5

e. What figures / tables do you anticipate producing?

state case trend and q1 to q2 income change
Historical trend and patterns of income
Comparison of effects across states - relate to states response?





